---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# eha

eha is a package for *event history and survival analysis*, with some 
enhancements of the *survival* package and many unique features, like *sampling
of risk sets* and the *weird bootstrap* in Cox regression, parametric proportional
hazards and accelerated failure time models allowing for right censoring and left
truncation. You can read more about that and other features in the `vignette("eha")`.


## Installation

You can install eha from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("goranbrostrom/eha")
```
If you are on Windows, you need to have
[Rtools](https://cran.r-project.org/bin/windows/Rtools/)
installed as well. Needed for compiling C and Fortran code. 
It is a good idea (but not necessary) to install
[MikTeX](http://www.miktex.org)
so you can get documentation in PDF format.  


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
## basic example code
```
